---
title: "Age-standardized synthetic-opioid death rates by sex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age-standardized synthetic-opioid death rates by sex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette turns multiple-cause-of-death (MCOD) records into an
age-standardized **synthetic-opioid** death rate by sex, for US residents only,
illustrated with 1999, 2010, and 2020. Synthetic opioids other than methadone are
ICD-10 **T40.4**, flagged by `flag_opioid_types()` as `other_synth_present`. T40.4
is the CDC-standard proxy for illicitly manufactured fentanyl, but it is a *proxy*
-- it also captures tramadol, U-47700, and other synthetics. Its composition shifts
over time: illicit fentanyl only came to dominate T40.4 around 2013-2014
(Ciccarone's "third wave"), so early-series T40.4 deaths are largely pharmaceutical
fentanyl and tramadol, not the illicit-fentanyl phenomenon of recent years. Read a
long T40.4 series as "synthetic opioids," not "fentanyl," throughout. Denominators
are the **SEER bridged-race** estimates (`race_scheme = "bridged"`), one internally
consistent race series, so the comparison is not confounded by a change in how the
Census classifies people by race.

All death data below are **small inline synthetic counts, clearly labelled
illustrative**; only the bundled population denominators are real. No chunk reads
a restricted record, an external file, or the network.

This vignette picks up where flagging leaves off. The *Classifying overdose
deaths* vignette (`vignette("classifying-overdose-deaths")`) builds the
`drug_death`/`opioid_death` flags used here and is the place to start if the flag
steps below are unfamiliar. For the full record-to-analysis workflow, begin with
`vignette("getting-started")`.

```{r libs}
library(narcan)
```

## From records to a fentanyl flag

Start with a handful of illustrative record-level rows. Real MCOD data carry
`restatus` (residency) and `ucod` (underlying cause) as raw columns, plus a
*derived* `f_records_all` -- the space-joined multiple-cause T-code string that
the flag helpers key on (`f_records_all` is produced by `unite_records()`; see
`vignette("getting-started")`).

```{r records}
records <- data.frame(
    year          = 2020L,
    restatus      = c(1L, 1L, 4L, 2L, 3L),
    ucod          = c("X42", "X44", "X42", "I250", "X42"),
    f_records_all = c("T404 T519", "T404", "T404", "I250", "T400"),
    age           = c(35L, 40L, 55L, 60L, 25L),
    sex           = c("male", "female", "male", "male", "female"),
    stringsAsFactors = FALSE
)
records
```

`subset_residents()` keeps `restatus %in% 1:3` and drops the column. The
non-resident row (`restatus == 4`) is removed.

```{r residents}
resid <- subset_residents(records)
resid
```

`flag_drug_deaths()` adds `drug_death` under the ISW7 (Injury Surveillance
Workgroup) rule -- a drug `ucod` **and** a T-code. The cardiac death (`I250`) is
not a drug death.

```{r drug}
resid <- flag_drug_deaths(resid, year = 2020)
resid[, c("ucod", "f_records_all", "drug_death")]
```

`flag_opioid_deaths()` adds `opioid_death`. `flag_opioid_types()` then splits the
opioid into type flags; `other_synth_present` is the T40.4 fentanyl proxy. Note
the T40.0-only row (opium, T40.0) is an opioid death but **not** a
synthetic-opioid death.

```{r opioid}
resid <- resid |>
    flag_opioid_deaths(year = 2020) |>
    flag_opioid_types(year = 2020)
resid[, c("f_records_all", "opioid_death", "other_synth_present")]
```

The fentanyl-proxy deaths are the rows with `other_synth_present == 1`:

```{r keep}
resid[resid$other_synth_present == 1, c("ucod", "f_records_all", "age", "sex")]
```

## Aggregate to counts, then rate

In practice you flag every record this way, then aggregate the fentanyl deaths to
`year x age x sex` counts before computing rates. We build that aggregated frame
directly here. **These counts are illustrative** -- a deterministic Poisson draw
tuned to rise steeply over time (the real synthetic-opioid surge) and to run
higher for males. `age` is the 5-year bin start (0, 5, ..., 85); `sex` is
`"male"`/`"female"`.

```{r counts}
counts <- expand.grid(
    year = c(1999L, 2010L, 2020L),
    age  = seq(0L, 85L, 5L),
    sex  = c("male", "female"),
    stringsAsFactors = FALSE
)
age_weight <- c(0, 0, .02, .2, .6, .9, 1, .95, .85, .7, .5, .3, .15, .07, .03,
                .01, 0, 0)
names(age_weight) <- seq(0, 85, 5)
year_factor <- c("1999" = 75, "2010" = 450, "2020" = 3000)   # ~40x rise
sex_factor  <- c("male" = 1.9, "female" = 1)

lambda <- year_factor[as.character(counts$year)] *
    sex_factor[counts$sex] *
    age_weight[as.character(counts$age)]
set.seed(42)
counts$deaths <- rpois(nrow(counts), lambda)
counts <- counts[order(counts$year, counts$sex, counts$age), ]
head(counts)
```

`add_pop_counts(race_scheme = "bridged")` joins the SEER bridged-race denominator
and adds `pop`. Race is not a `by_var`, so `pop` is the all-race total for each
`year x age x sex` cell; bridged **requires** `year` in `by_vars`. With no
`state_fips`/`county_fips`, the national denominator is used automatically.

```{r pop}
counts <- add_pop_counts(counts, by_vars = c("year", "age", "sex"),
                         race_scheme = "bridged")
head(counts[, c("year", "age", "sex", "deaths", "pop")])
```

`add_std_pop()` adds the US 2000 standard population (`pop_std`) and its unit
weights (`unit_w`, summing to 1 across the 18 five-year age groups). `s204` is
narcan's code for that standard population in 18 five-year age bins (the default
`std_cat`); see `?add_std_pop` for single-year alternatives, which must match
your age binning. Other standard populations are available too (e.g., the Segi
world standard and alternative age binnings) -- see `?add_std_pop` and the
`std_pops` dataset for the full list of `std_cat` codes.

```{r std}
counts <- add_std_pop(counts, std_cat = "s204", by_vars = "age")
head(counts[, c("age", "pop_std", "unit_w")])
```

`calc_asrate_var()` adds the **age-specific** rate (`fentanyl_rate`, per 100,000)
and its Poisson variance (`fentanyl_var`) for each age-sex-year cell -- the rate
*before* standardization. Here are the 2020 male age-specific rates, which peak in
early adulthood (~age 30) and decline thereafter:

```{r asrate}
counts <- calc_asrate_var(counts, new_name = fentanyl, death_col = deaths,
                          pop_col = pop)
subset(counts, year == 2020L & sex == "male",
       select = c("age", "deaths", "pop", "fentanyl_rate"))
```

`calc_stdrate_var()` collapses the age bins into one age-standardized rate per
`year x sex`, reweighting each stratum to the US 2000 age structure. Pass the
grouping columns (`year`, `sex`) explicitly -- they are not added automatically.

```{r stdrate}
std <- calc_stdrate_var(counts, fentanyl_rate, fentanyl_var, year, sex)
as.data.frame(std)
```

## Reading the result

In these illustrative data the male rate runs roughly twice the female rate in
every year, and both rise about 35-40 fold from 1999 to 2020 -- the shape of the
real synthetic-opioid epidemic, if not the exact levels. A 95% confidence
interval follows from the returned variance as
`fentanyl_rate +/- 1.96 * sqrt(fentanyl_var)`:

```{r ci}
std <- as.data.frame(std)
std$lower <- std$fentanyl_rate - 1.96 * sqrt(std$fentanyl_var)
std$upper <- std$fentanyl_rate + 1.96 * sqrt(std$fentanyl_var)
std[, c("year", "sex", "fentanyl_rate", "lower", "upper")]
```

This normal-approximation (Wald) interval is fine for well-populated cells but
under-covers when death counts are small -- it can even return a negative lower
bound. For sparse strata (sub-national rates, or fine demographic cells; NCHS
treats a rate from fewer than 20 deaths as unreliable), use a gamma-based interval
(Fay and Feuer, *Stat Med* 1997) rather than this Wald form.

Because both sexes are standardized to the same US 2000 age structure, the
male-female gap reflects differences in death *rates*, not in age composition --
that is the point of age standardization.

Plotting `std` makes both patterns visible at once -- the steep rise over time and
the persistent male-female gap. The chunk below needs ggplot2 (a Suggests-only
dependency), so it only runs when that package is installed.

```{r plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.alt = "Line plot of age-standardized synthetic-opioid death rate per 100,000 against year for 1999, 2010, and 2020, with separate lines for males and females. Both lines rise steeply across the period, and the male line sits roughly twice as high as the female line in every year. Vertical bars show 95% confidence intervals.", fig.cap = "Age-standardized synthetic-opioid (fentanyl-proxy) death rate per 100,000 by sex, illustrative data. Points are standardized rates and bars are 95% confidence intervals."}
library(ggplot2)

ggplot(std, aes(x = year, y = fentanyl_rate, color = sex)) +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 1.2) +
    geom_line() +
    geom_point(size = 2) +
    scale_x_continuous(breaks = unique(std$year)) +
    labs(x = "Year", y = "Age-standardized rate per 100,000",
         color = "Sex") +
    theme_minimal()
```

These counts are synthetic and illustrative only. A real analysis flags the
restricted or public MCOD records exactly as shown above (see
`vignette("classifying-overdose-deaths")`), aggregates them, and runs this same
rate pipeline.

## See also

- `vignette("getting-started")` -- the full pipeline and where rates fit.
- `vignette("population-denominators")` -- choosing and joining the denominator.
- `vignette("hispanic-origin")` -- add a Hispanic-origin stratum to these rates.
- `vignette("geography-fips")` -- compute the same rates at sub-national grains.
