---
title: "Age-standardized fentanyl death rates by sex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age-standardized fentanyl death rates by sex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette turns multiple-cause-of-death (MCOD) records into an
age-standardized synthetic-opioid death rate by sex, for US residents only,
across 1999-2024. "Fentanyl" here means synthetic opioids other than methadone
(ICD-10 **T40.4**), flagged by `flag_opioid_types()` as `other_synth_present`.
T40.4 is the CDC-standard fentanyl proxy, but it is a *proxy* -- it also captures
tramadol, U-47700, and other synthetics, so a small share of these deaths are not
fentanyl. Denominators are the **SEER bridged-race** estimates
(`race_scheme = "bridged"`, 1969-2024), one era-consistent series that predates
the fentanyl surge, so a 1999-vs-2024 comparison is not confounded by a change in
how the Census counts people.

All death data below are **small inline synthetic counts, clearly labelled
illustrative**; only the bundled population denominators are real. No chunk reads
a restricted record, an external file, or the network.

This vignette picks up where flagging leaves off. The *Classifying overdose
deaths* vignette builds the `drug_death`/`opioid_death` flags used here and is the
place to start if the flag steps below are unfamiliar.

```{r libs}
library(narcan)
```

## From records to a fentanyl flag

Start with a handful of illustrative record-level rows. Real MCOD data carry
`restatus` (residency), `ucod` (underlying cause), and `f_records_all` (the
space-joined multiple-cause T-code string) that the flag helpers key on.

```{r records}
records <- data.frame(
    year          = 2020L,
    restatus      = c(1L, 1L, 4L, 2L, 3L),
    ucod          = c("X42", "X44", "X42", "I250", "X42"),
    f_records_all = c("T404 T519", "T404", "T404", "I250", "T400"),
    age           = c(35L, 40L, 55L, 60L, 25L),
    sex           = c("male", "female", "male", "male", "female"),
    stringsAsFactors = FALSE
)
records
```

`subset_residents()` keeps `restatus %in% 1:3` and drops the column. The
non-resident row (`restatus == 4`) is removed.

```{r residents}
resid <- subset_residents(records)
resid
```

`flag_drug_deaths()` adds `drug_death` under the ISW7 (Injury Surveillance
Workgroup) rule -- a drug `ucod` **and** a T-code. The cardiac death (`I250`) is
not a drug death.

```{r drug}
resid <- flag_drug_deaths(resid, year = 2020)
resid[, c("ucod", "f_records_all", "drug_death")]
```

`flag_opioid_deaths()` adds `opioid_death`. `flag_opioid_types()` then splits the
opioid into type flags; `other_synth_present` is the T40.4 fentanyl proxy. Note
the T40.0-only row (opium, T40.0) is an opioid death but **not** a
synthetic-opioid death.

```{r opioid}
resid <- resid |>
    flag_opioid_deaths(year = 2020) |>
    flag_opioid_types(year = 2020)
resid[, c("f_records_all", "opioid_death", "other_synth_present")]
```

The fentanyl-proxy deaths are the rows with `other_synth_present == 1`:

```{r keep}
resid[resid$other_synth_present == 1, c("ucod", "f_records_all", "age", "sex")]
```

## Aggregate to counts, then rate

In practice you flag every record this way, then aggregate the fentanyl deaths to
`year x age x sex` counts before computing rates. We build that aggregated frame
directly here. **These counts are illustrative** -- a deterministic Poisson draw
tuned to rise steeply over time (the real synthetic-opioid surge) and to run
higher for males. `age` is the 5-year bin start (0, 5, ..., 85); `sex` is
`"male"`/`"female"`.

```{r counts}
counts <- expand.grid(
    year = c(1999L, 2010L, 2020L),
    age  = seq(0L, 85L, 5L),
    sex  = c("male", "female"),
    stringsAsFactors = FALSE
)
age_weight <- c(0, 0, .02, .2, .6, .9, 1, .95, .85, .7, .5, .3, .15, .07, .03,
                .01, 0, 0)
names(age_weight) <- seq(0, 85, 5)
year_factor <- c("1999" = 75, "2010" = 450, "2020" = 3000)   # ~40x rise
sex_factor  <- c("male" = 1.9, "female" = 1)

lambda <- year_factor[as.character(counts$year)] *
    sex_factor[counts$sex] *
    age_weight[as.character(counts$age)]
set.seed(42)
counts$deaths <- rpois(nrow(counts), lambda)
counts <- counts[order(counts$year, counts$sex, counts$age), ]
head(counts)
```

`add_pop_counts(race_scheme = "bridged")` joins the SEER bridged-race denominator
and adds `pop`. Race is not a `by_var`, so `pop` is the all-race total for each
`year x age x sex` cell; bridged **requires** `year` in `by_vars`. With no
`state_fips`/`county_fips`, the national denominator is used automatically.

```{r pop}
counts <- add_pop_counts(counts, by_vars = c("year", "age", "sex"),
                         race_scheme = "bridged")
head(counts[, c("year", "age", "sex", "deaths", "pop")])
```

`add_std_pop()` adds the US 2000 standard population (`pop_std`) and its unit
weights (`unit_w`, summing to 1 across the 18 five-year age groups). `s204` is
narcan's code for that standard population in 18 five-year age bins (the default
`std_cat`); see `?add_std_pop` for single-year alternatives, which must match
your age binning.

```{r std}
counts <- add_std_pop(counts, std_cat = "s204", by_vars = "age")
head(counts[, c("age", "pop_std", "unit_w")])
```

`calc_asrate_var()` adds the **age-specific** rate (`fentanyl_rate`, per 100,000)
and its Poisson variance (`fentanyl_var`) for each age-sex-year cell -- the rate
*before* standardization. Here are the 2020 male age-specific rates, which peak in
early adulthood (~age 30) and decline thereafter:

```{r asrate}
counts <- calc_asrate_var(counts, new_name = fentanyl, death_col = deaths,
                          pop_col = pop)
subset(counts, year == 2020L & sex == "male",
       select = c("age", "deaths", "pop", "fentanyl_rate"))
```

`calc_stdrate_var()` collapses the age bins into one age-standardized rate per
`year x sex`, reweighting each stratum to the US 2000 age structure. Pass the
grouping columns (`year`, `sex`) explicitly -- they are not added automatically.

```{r stdrate}
std <- calc_stdrate_var(counts, fentanyl_rate, fentanyl_var, year, sex)
as.data.frame(std)
```

## Reading the result

In these illustrative data the male rate runs roughly twice the female rate in
every year, and both rise about 35-40 fold from 1999 to 2020 -- the shape of the
real synthetic-opioid epidemic, if not the exact levels. A 95% confidence
interval follows from the returned variance as
`fentanyl_rate +/- 1.96 * sqrt(fentanyl_var)`:

```{r ci}
std <- as.data.frame(std)
std$lower <- std$fentanyl_rate - 1.96 * sqrt(std$fentanyl_var)
std$upper <- std$fentanyl_rate + 1.96 * sqrt(std$fentanyl_var)
std[, c("year", "sex", "fentanyl_rate", "lower", "upper")]
```

Because both sexes are standardized to the same US 2000 age structure, the
male-female gap reflects differences in death *rates*, not in age composition --
that is the point of age standardization.

These counts are synthetic and illustrative only. A real analysis flags the
restricted or public MCOD records exactly as shown above (see the "Classifying
overdose deaths" vignette), aggregates them, and runs this same rate pipeline.
