---
title: "Classifying overdose deaths (ISW7)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classifying overdose deaths (ISW7)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

narcan operationalizes the Injury Surveillance Workgroup (ISW7; CSTE/CDC) case
definitions for drug- and opioid-involved overdose deaths, reading them straight
off ICD-10 multiple-cause-of-death (MCOD) records. A death is classified from two
fields -- the **underlying cause** (`ucod`, e.g. `"X42"`) and the space-joined
string of all contributory **T-codes** (`f_records_all`, e.g. `"T404 T406"`). One
caveat matters for fentanyl. T40.4 ("synthetic opioids other than methadone") is
the CDC-standard proxy for illicitly manufactured fentanyl, which has dominated
that code since roughly 2013 -- but narcan carries no fentanyl-*only* flag,
because the ICD-10 code itself cannot separate fentanyl from other synthetic
opioids.

```{r libs}
library(narcan)
```

## An illustrative record-level frame

The rows below are **synthetic and illustrative** -- hand-written to span the
opioid subtypes, not real NCHS records. Each row carries only the two fields the
classifiers key on, plus `restatus` (residency status). Six rows walk through the
opioid subtypes, one is a non-opioid drug death, two are non-drug deaths, and one
is a non-resident (`restatus = 4`).

```{r frame}
deaths <- data.frame(
    ucod = c(
        "X42", "X42", "X44", "X42", "X42",
        "Y12", "X40", "I250", "C509", "X42"
    ),
    f_records_all = c(
        "T400", "T401 T404", "T402", "T403", "T404",
        "T406", "T436", "I250", "C509", "T404 T401"
    ),
    restatus = c(1L, 1L, 1L, 3L, 1L, 2L, 1L, 1L, 1L, 4L),
    stringsAsFactors = FALSE
)
deaths
```

## Keep US residents only

`subset_residents()` keeps `restatus %in% 1:3` (US residents) and drops the
`restatus` column once it has filtered on it.

```{r residents}
resident_deaths <- subset_residents(deaths)
nrow(deaths)            # 10 synthetic rows
nrow(resident_deaths)   # non-resident (restatus 4) dropped -> 9
```

## Run the flag pipeline, one step at a time

Three flaggers run in sequence. Each one adds columns and never drops a row, so
you can watch the classification build up step by step. Pass `year` to select the
coding era.

In real use, `f_records_all` is not hand-built -- you read the raw fixed-width
MCOD file with `import_mcod_fwf()` and collapse its multiple-cause fields with
`unite_records()`. See `vignette("getting-started")` for that on-ramp, which is
where a newcomer with real data should start.

**Step 1.** `flag_drug_deaths()` adds `drug_death`. The two non-drug rows (I250,
C509) score 0; every drug-poisoning row scores 1, including the psychostimulant
row (e.g. methamphetamine) that is not an opioid.

```{r step1}
step1 <- flag_drug_deaths(resident_deaths, year = 2019L)
step1[, c("ucod", "f_records_all", "drug_death")]
```

**Step 2.** `flag_opioid_deaths()` adds `opioid_death`. It lights up only where a
drug death *also* carries an opioid T-code, so `X40` / `T436` stays 0.

```{r step2}
step2 <- flag_opioid_deaths(step1, year = 2019L)
step2[, c("ucod", "f_records_all", "drug_death", "opioid_death")]
```

**Step 3.** `flag_opioid_types()` adds the six subtype indicators plus
`num_opioids` (and `multi_opioids`). The polydrug row (`T401 T404`) now reads two
distinct opioids.

```{r step3}
classified <- flag_opioid_types(step2, year = 2019L)
classified[, c("f_records_all", "opioid_death", "heroin_present",
               "other_synth_present", "num_opioids", "multi_opioids")]
```

## The two ISW7 rules

The definitions nest. A **drug death** requires a drug-poisoning underlying cause
(X40-44, X60-64, X85, or Y10-14) *and* a drug T-code (T36.0-T50.9). An **opioid
death** is the strict subset of drug deaths that also carry an opioid T-code
(T40.0-T40.4 or T40.6). So a poisoning with a non-opioid drug is a drug death but
not an opioid death. Row `X40` / `T436` (a psychostimulant, e.g. methamphetamine)
is exactly that edge case -- `drug_death` is 1, `opioid_death` is 0.

```{r edge}
classified[classified$ucod == "X40",
           c("ucod", "f_records_all", "drug_death", "opioid_death")]
```

## The opioid subtypes

Among opioid deaths, `flag_opioid_types()` breaks out six specific T40 subtypes --
`opium_present` (T40.0), `heroin_present` (T40.1), `other_natural_present`
(T40.2), `methadone_present` (T40.3), `other_synth_present` (T40.4, the fentanyl
proxy), and `other_op_present` (T40.6). `unspecified_op_present` is the residual for an opioid
death where none of the six matched; for ICD-10 data (1999+) it is **always 0 by
construction**, because every ICD-10 opioid T40 code falls into at least one of
those six subtypes -- it fires only for the pre-1999 ICD-9 residual (code 965.0).
If you want the share of opioid deaths with an *unspecified* opioid type in modern
data, the column you want is `other_op_present` (T40.6, "other and unspecified
narcotics"), not the similarly named `unspecified_op_present`; see
`vignette("unspecified-drug-deaths")`. `num_opioids` counts how many distinct
subtypes appear on the record, and `multi_opioids` is 1 when `num_opioids > 1`.
The polydrug row (`T401 T404`) shows `num_opioids = 2`.

```{r subtypes}
subtype_cols <- c(
    "opium_present", "heroin_present", "other_natural_present",
    "methadone_present", "other_synth_present", "other_op_present",
    "unspecified_op_present", "num_opioids"
)
classified[classified$opioid_death == 1, c("f_records_all", subtype_cols)]
```

## Matching `year` to the coding era

Every flagger dispatches on `year`. Data years before 1999 use ICD-9 logic; 1999
onward uses the ICD-10 rules shown here. Match `year` to the data year of the
records so the correct code list is applied -- there is no default era.

Counts are **not directly comparable across the 1999 ICD-9-to-ICD-10 revision**.
NCHS documents that drug-poisoning death counts before and after the revision
require a comparability-ratio adjustment before they can be joined into one trend
-- the numerator analog of the "not comparable" caveat on the denominator schemes
(see `vignette("population-denominators")`). Concatenating raw pre- and post-1999
counts produces an artifactual step at the boundary.

## See also

- `vignette("population-denominators")` -- attach a matching population to these
  flagged counts, the next step before computing any rate.
- `vignette("getting-started")` -- the full raw-data-to-rate pipeline and where
  this vignette fits.
- Related flags not covered here: `flag_opioid_contributed()` (opioids as a
  contributory rather than underlying cause), `flag_nonopioid_drug_deaths()`, and
  `flag_od_intent()` (unintentional / suicide / homicide / undetermined intent).
