---
title: "Demographic recodes across coding eras"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demographic recodes across coding eras}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

NCHS changed how it codes age, sex, and race several times over the decades. narcan's
recoders normalize each raw field to one stable scheme, dispatching on the data year.
The recurring hazard is that codes which look continuous across a revision are not,
and chaining them produces artifactual trends. This vignette walks the three families
-- age, sex, and race -- and shows where each one breaks.

Every input below is a small illustrative vector, not an NCHS record, and every chunk
runs.

```{r libs}
library(narcan)
```

## Age: two recodes, two outputs

narcan carries two age recoders because MCOD files carry two different age fields.

**Detail age (`age`) to completed years.** `remap_age()` reads the unit-coded
detail-age field and returns completed years in a new `age_years` column. The encoding
changed at 2003 (a 3-digit code before, 4-digit after), so pass `year`. Sub-year ages
(months, weeks, days, hours, minutes) collapse to 0; not-stated codes become `NA`.

```{r remap-age}
remap_age(data.frame(year = 2000, age = c(37, 206, 999)))$age_years    # pre-2003
remap_age(data.frame(year = 2019, age = c(1037, 2006, 1999, 9999)))$age_years  # 2003+
```

Pre-2003, a years-unit code is the age itself (37 -> 37) and `206` is 6 months -> 0.
From 2003 a leading `1` marks years, so `1037` -> 37 and `2006` (months) -> 0; both
`1999` and `9999` are not-stated -> `NA`.

**AGER27 recode (`ager27`) to 5-year bins.** `convert_ager27()` maps the NCHS
27-level age recode (codes 1-26 are age groups, 27 is "not stated" -> `NA`) to 5-year-bin
starts (0, 5, ..., 85). The finer under-5 categories collapse into the 0-4 bin, and
everything 85 and older into 85.

```{r ager27}
convert_ager27(data.frame(ager27 = c(1, 7, 10, 23, 27)))
```

**Label the bins.** `categorize_age_5()` turns 5-year-bin starts into an ordered
factor with readable labels -- the form `tidyr::complete()` and ggplot2 expect.

```{r cat-age}
categorize_age_5(seq(0, 85, 5))
```

The two age outputs are **not** interchangeable. `age_years` is single completed
years; the AGER27 path -- and the standard populations `add_std_pop()` uses -- are
5-year bins. To bin `age_years` yourself:

```{r bin-years}
age_years <- c(0, 3, 37, 88)
pmin(floor(age_years / 5) * 5, 85)     # -> 0, 0, 35, 85
```

## Sex: numeric before 2003, character after

`categorize_sex()` maps either coding -- numeric `1`/`2` (1979-2002) or character
`"M"`/`"F"` (2003+) -- to `"male"`/`"female"`, the labels `add_pop_counts()` joins on.
Pass `year`; it is authoritative.

```{r sex}
categorize_sex(c(1, 2, 9), year = 2000)       # pre-2003 numeric
categorize_sex(c("M", "F", "U"), year = 2019) # 2003+ character
categorize_female(c(1, 2), year = 2000)       # -> 0 1 (male = 0, female = 1)
```

The trap: omit `year` and the era is guessed from the column type. A pre-2003 sex
column re-read from a CSV as the **characters** `"1"`/`"2"` is guessed to be the
modern `"M"`/`"F"` scheme and maps to all-`NA` -- loudly, with warnings.

```{r sex-trap}
categorize_sex(c("1", "2"), year = NULL)     # guesses modern -> NA (with warnings)
categorize_sex(c("1", "2"), year = 2000)     # year fixes it -> male, female
```

## Race: the most-revised field

Race coding changed repeatedly, and `remap_race()` dispatches on `year` to a
standardized code that `categorize_race()` then labels. These eras are genuinely
different schemes, not one evolving list.

**1992-2020 (bridged race).** Detailed codes 1-7, plus a family of "other" codes that
all collapse to 99.

```{r race-bridged}
r <- remap_race(data.frame(year = 2004, race = c(1, 2, 3, 6, 18)), year = 2004)
r$race                          # 1 2 3 6 99
categorize_race(r$race)         # white, black, american_indian, hawaiian, other
```

The Asian and Pacific Islander subgroups (Chinese, Japanese, Hawaiian, Filipino) are
resolved separately in this era but have no matching bridged population denominator,
so collapse them to a single API group before computing rates.

**Earlier eras differ.** The same raw code means different things before 1992 -- and
there are two distinct pre-1992 schemes (1979-1988 and 1989-1991), not one. Code 8, for
instance, is Filipino here (1979-1988, mapped to 7) but "other" (99) in 1989-1991. This
is exactly why `remap_race()` refuses to run without a year.

```{r race-early}
remap_race(data.frame(year = 1985, race = c(0, 1, 7, 8)), year = 1985)$race  # 99 1 99 7
```

**2021 is a gap.** The bridged race fields were dropped and the single-race recodes
were not yet populated, so `remap_race()` sets race to `NA` (with a warning).

```{r race-2021}
remap_race(data.frame(year = 2021, race = c(1, 2)), year = 2021)$race
```

**2022+ (single race).** From 2022 the bridged column is gone; `remap_race()` reads
the single-race Race Recode (`racer5`) and maps it into a **non-colliding** code space
(101-106) so the two schemes cannot be silently mixed. `categorize_race()` labels
these with an `_only` suffix (except code 106, `multiracial`).

```{r race-single}
r22 <- remap_race(data.frame(year = 2022, racer5 = c(1, 2, 4, 6)), year = 2022)
r22$race                                # 101 102 104 106
categorize_race(c(101, 102, 104, 106))  # white_only, black_only, asian_only, multiracial
```

Both `remap_race()` and `categorize_race()` warn on the single-race codes, because
bridged (2020 and earlier) and single-race (2022+) categories are **not comparable**.
Do not chain them into one trend.

## Why the year matters

Two coding boundaries hide in these recoders, and neither is the ICD boundary:

- **Age and sex** change their coding at **2003** (the death-certificate revision).
- **Race** changes repeatedly, with the sharpest break at **2021/2022** (bridged to
  single race).

Neither lines up with the ICD-9-to-ICD-10 boundary at **1999** that the flag functions
dispatch on. A long multi-year analysis therefore carries several independent
comparability breaks at once, so each recoder reads the data year to apply the right
map, and not-stated codes always become `NA` rather than a guessed value.

These recoders produce exactly the `age`, `sex`, and `race` columns that
`add_pop_counts()` joins on, with labels that match the population data.

## See also

- `vignette("getting-started")` -- where these recodes sit in the full pipeline.
- `vignette("population-denominators")` -- the denominators these columns join to,
  and the matching race-scheme caveats.
- `vignette("real-data-end-to-end")` -- the recoders applied to a real public-use file.
