---
title: "Harmonizing geography with FIPS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Harmonizing geography with FIPS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

County identifiers are not stable across data years. New counties appear, a few
merge or are renamed, and -- more disruptively -- NCHS does not encode state the
same way in every era. Public multiple-cause-of-death (MCOD) files use NCHS
numeric state codes through 2002 and 2-letter postal abbreviations from 2003 on,
and the numeric schemes overlap while meaning different states (NCHS `"06"` is
Colorado; FIPS `"06"` is California). `add_county_fips()` translates whichever
scheme a file uses into one 5-digit `county_fips` that is comparable across
years.

**Public-data caveat.** Public-use MCOD carries state and county geography only
through **2004**. From 2005 on, any sub-national analysis requires the restricted
NCHS All-County files. Every example below therefore uses data years
**1999-2004**, and nothing here works for later years without the restricted data.

All data here are small synthetic frames plus bundled narcan crosswalks
(`st_fips_map`, `ihme_fips`); no chunk downloads anything and no restricted NCHS
records are used.

```{r libs}
library(narcan)
```

## `state_abbrev_to_fips()` -- one column at a time

The simplest helper maps a postal abbreviation to its zero-padded 2-digit state
FIPS. Pair the input with the output to see the mapping:

```{r abbrev}
abbrevs <- c("CA", "NY", "TX", "PA")
data.frame(abbrev = abbrevs, state_fips = state_abbrev_to_fips(abbrevs))
```

`"CA"` becomes `"06"`, not `"6"` -- the zero-padding is what lets the result feed
a FIPS-keyed join. Unrecognized abbreviations and the foreign/unknown code `"ZZ"`
return `NA` (with a warning) rather than passing through unconverted.

## `add_county_fips()` -- the core call

Feed `add_county_fips()` the raw county column as **character** (`countyrs` for
county of residence, `countyoc` for occurrence). A numeric column is refused,
because numeric input has already dropped the leading zero (`01001` reads as
`1001`, whose first two digits parse as a different state). Pass `year` -- a
scalar or a `year` column -- so the coding scheme is chosen deterministically.

Start with a small synthetic frame that straddles the 2002/2003 boundary. The
1999-2002 rows carry NCHS numeric state codes; the 2003-2004 rows carry postal
abbreviations, exactly as the public files do:

```{r deaths_frame}
deaths <- tibble::tibble(
    id       = 1:5,
    year     = c(1999L, 2001L, 2002L, 2003L, 2004L),
    countyrs = c("06031", "36061", "48059", "CA075", "TX201")
)
deaths
```

Now harmonize. The call adds two columns -- `st_fips` (2-digit state) and
`county_fips` (5-digit state + county) -- decoding each row under its own era's
scheme:

```{r add_fips}
harmonized <- add_county_fips(deaths, countyrs)
harmonized[c("id", "year", "countyrs", "st_fips", "county_fips")]
```

Read the before/after together. The raw `countyrs` string `"06031"` (1999) maps
to `st_fips` `"08"` and `county_fips` `"08031"` -- NCHS `"06"` is **Colorado**,
not California. In 2003, California enters instead through the abbreviation
`"CA"`, giving `"06075"`. A single numeric code like `"06..."` is therefore
ambiguous on its own; `add_county_fips()` resolves it row by row from the year,
so one frame spanning both eras comes out on a consistent 5-digit FIPS grain.
Codes it cannot place -- `"ZZ"` foreign residence, or the ambiguous NCHS code
`"62"` -- yield `NA` rather than a spurious string.

## The bundled crosswalks

Both helpers read `st_fips_map`, which lines up state name, postal abbreviation,
FIPS code, and NCHS code so the two numeric schemes can be told apart:

```{r st_fips_map}
head(st_fips_map)
```

`ihme_fips` handles the other half of the problem -- counties that split, merge,
or are renamed over a long series. It collapses unstable `orig_fips` codes to a
temporally stable `ihme_fips` (from Dwyer-Lindgren et al., JAMA 2016), so a
county can be compared across periods:

```{r ihme_fips}
head(ihme_fips)
```

For a multi-year county analysis, remap `county_fips` through `ihme_fips` before
comparing -- several Alaska areas and Virginia independent cities are otherwise
not comparable across the 1999-2004 window.

## From FIPS to denominators

Once `county_fips` or a `state_fips` column exists, it is the join key for
sub-national denominators. `add_county_fips()` names its state column `st_fips`;
`add_pop_counts()` keys on `state_fips`, so rename `st_fips` to `state_fips` for a
state join, or use the 5-digit `county_fips` directly for a county join (see the
companion `vignette("population-denominators")`). As above, this path is only
valid for **1999-2004** public data -- 2005+ sub-national denominators require
the restricted NCHS All-County files.

## See also

- `vignette("population-denominators")` -- sub-national denominators use these
  FIPS-harmonized geographies as their join key.
- `vignette("getting-started")` -- the package overview and where this vignette
  fits in the pipeline.
