---
title: "Getting started with narcan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with narcan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

narcan turns raw US multiple-cause-of-death (MCOD) micro-data into drug- and
opioid-overdose death counts and age-standardized rates. It implements the Injury
Surveillance Workgroup (ISW7; CSTE/CDC) case definitions across both the ICD-9
(1979-1998) and ICD-10 (1999+) eras, and harmonizes the race, Hispanic-origin, and
county recodes that NCHS changed over the years. This vignette maps the whole
package; each step links to a dedicated vignette.

```{r libs}
library(narcan)
```

## The pipeline

Most analyses follow the same four steps, from a raw NCHS file to a rate.

| Step | Functions | Vignette |
|------|-----------|----------|
| 1. Import | `import_mcod_fwf()`, `unite_records()` | this vignette (below) |
| 2. Flag | `flag_drug_deaths()`, `flag_opioid_deaths()`, `flag_opioid_types()` | `vignette("classifying-overdose-deaths")` |
| 3. Denominators | `add_pop_counts()` | `vignette("population-denominators")` |
| 4. Rates | `add_std_pop()`, `calc_stdrate_var()` | `vignette("age-standardized-rates")` |

Further vignettes branch off this spine: `vignette("demographic-recodes")` (the
year-aware age/sex/race recodes that produce the columns above),
`vignette("hispanic-origin")` (Hispanic-origin stratification), and
`vignette("geography-fips")` (sub-national geography). A measurement caveat --
how the share of *unspecified* drug deaths shifts trends over time -- is in
`vignette("unspecified-drug-deaths")`. To see every step run on a real file,
`vignette("real-data-end-to-end")` takes the public-use 2004 data all the way to an
age-standardized opioid rate.

## Getting the data

Public-use MCOD micro-data is distributed by NCHS and mirrored by NBER; the
restricted All-County files require an NCHS data-use agreement (DUA). narcan can
fetch a public-use year for you:

```{r download, eval = FALSE}
# Downloads the NBER public-use mirror into `download_dir`.
download_mcod_csv(2019, download_dir = "raw_data")   # or download_mcod_dta()
```

To read a raw fixed-width file (public or restricted), use `import_mcod_fwf()`,
which applies narcan's byte-verified column dictionary for the given year and
tier:

```{r import, eval = FALSE}
raw <- import_mcod_fwf("mort2019us.dat", year = 2019, tier = "public")
```

The two tiers share the within-record layout; the public tier blanks sub-state
geography (from 2005) and a few certifier-entered items, which `import_mcod_fwf()`
returns as all-`NA` so public and restricted output are column-compatible.

**No restricted-data agreement yet?** The public-use file for **2004** is the last
public year that still carries county geography (2005 onward suppress it).
`import_mcod_fwf()` orders its output to the restricted column layout for every
tier, so reading the 2004 public file already gives you a restricted-shaped frame
with county of residence and occurrence -- a DUA-free stand-in for developing
county-level code before you obtain the restricted All-County files:

```{r public-standin, eval = FALSE}
raw04 <- import_mcod_fwf("mort2004us.dat", year = 2004, tier = "public")
```

Two caveats. On the public file, county FIPS are populated only for counties with a
population of at least 100,000; smaller (disproportionately rural)
counties collapse to a residual code, so this stand-in exercises large-county
behavior only. And `tier = "restricted"` adds nothing for 2004 -- the single field
it would add (`racer40`) sits past the end of the public record and returns
all-`NA`.

## From raw records to the flag pipeline

The flaggers key on two fields: the underlying cause (`ucod`) and a single string
of all contributory cause codes (`f_records_all`). A real MCOD file stores the
contributory codes across many separate `record_*` columns, so `f_records_all` is
**derived**, not read directly -- `unite_records()` builds it:

```{r unite, eval = FALSE}
records <- unite_records(raw, year = 2019)

# `records` now carries `ucod` + the collapsed `f_records_all`, ready to flag:
flagged <- records |>
    flag_drug_deaths(year = 2019) |>
    flag_opioid_deaths(year = 2019)
```

From here you are at the start of `vignette("classifying-overdose-deaths")`, which
walks the flag pipeline step by step. To run unite + the drug/opioid/type/intent
flaggers in a single call, use the `flag_all_deaths()` convenience wrapper.

## The columns narcan expects

Downstream functions read a handful of raw (or lightly recoded) MCOD columns. This
is the minimal set; each vignette introduces the ones it needs.

| Column | Meaning | Raw or derived |
|--------|---------|----------------|
| `year` | data year (4-digit; 1979-1995 files use a 2-digit `datayear`) | raw |
| `ucod` | underlying cause of death (ICD code) | raw |
| `record_1`, `record_2`, ... | contributory cause-code columns | raw |
| `f_records_all` | space-joined contributory codes | derived (`unite_records()`) |
| `restatus` | residency status (1-3 = US resident) | raw |
| `age` | age (see `remap_age()` / `categorize_age_5()`) | raw / recoded |
| `sex` | sex (see `categorize_sex()`) | raw / recoded |
| `race` | race (see `categorize_race()`; era-dependent) | raw / recoded |
| `hspanicr` | Hispanic-origin recode (see `categorize_hispanic_origin()`) | raw |
| `countyrs` / `countyoc` | county of residence / occurrence | raw |

## Next steps

Start with `vignette("classifying-overdose-deaths")` (the flag step), then
`vignette("population-denominators")` and `vignette("age-standardized-rates")` to
turn counts into rates. `vignette("real-data-end-to-end")` runs that whole path on a
real public-use file. Branch to `vignette("demographic-recodes")` for the age/sex/race
recoders, `vignette("hispanic-origin")` to stratify by Hispanic origin, or
`vignette("geography-fips")` for sub-national work;
`vignette("unspecified-drug-deaths")` covers a measurement caveat.

## A note on this documentation

Every runnable chunk in these vignettes uses small **synthetic or bundled** data,
never restricted records, so they build without a data-use agreement or a network
connection. The download/import chunks above are shown with `eval = FALSE` because
they need a real file; swap in your own path to run them.
