---
title: "Hispanic origin: the two recode pairs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hispanic origin: the two recode pairs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

narcan gives you two ways to recode the NCHS `hspanicr` (Hispanic Origin/Race
Recode) field, and picking the wrong one is a common, quiet mistake. This
vignette is about which pair to use, and why they are not interchangeable.

All data here are synthetic death counts joined to bundled public Census/SEER
population estimates; no restricted NCHS records are used, and no chunk downloads
anything.

```{r libs}
library(narcan)
```

## Two pairs, two jobs

| | Detailed pair | Binary pair |
|---|---|---|
| Vectorized recode | `categorize_hspanicr()` | `categorize_hispanic_origin()` |
| Add-a-column helper | `add_hspanicr_column()` | `add_hispanic_origin()` |
| Output | 9- or 14-category ethnicity factor | `"hispanic"` / `"non_hispanic"` / `"unknown"` |
| Use for | descriptive counts and proportions by subgroup | **rates** (it has a matching denominator) |

The **detailed** pair returns the full NCHS recode -- Mexican, Puerto Rican,
Cuban, and so on, with non-Hispanic race detail. It is the right tool for a
table or figure that breaks Hispanic deaths into subgroups. It has **no matching
population denominator**: Census and SEER do not publish population by these
subgroups, so you cannot turn these counts into rates.

The **binary** pair collapses the same field to the only Hispanic-origin
granularity the population data resolve -- Hispanic vs non-Hispanic. That is the
recode to use when you are going to divide by a population.

```{r both}
codes <- c(1, 5, 6, 9)   # 9-category: Mexican, Other/unknown Hispanic,
                         # non-Hispanic white, Hispanic origin unknown
categorize_hspanicr(codes, year = 2019)          # detailed (subgroups)
categorize_hispanic_origin(codes, year = 2019)   # binary (for rates)
```

`categorize_hspanicr()` prints as an ordered factor (`9 Levels: mexican <
puerto_rican < cuban < ...`); the `<` is just R's print order for a canonical
tabulation sequence, not a ranking.

Two subtleties are baked into the binary recode. Code 5 ("Other or unknown
Hispanic") **is** Hispanic, so it maps to `"hispanic"`; only code 9 ("Hispanic
origin unknown/not stated") maps to `"unknown"`. And `year` is required, because
the recode changed from 9 categories (1989-2020) to 14 (2022+) and is reserved
(not populated) in 2021 -- the two schemes are not comparable, though the binary
axis they collapse to is.

## Adding the column and joining a denominator

`add_hispanic_origin()` reads the data year per row (from `year`, or two-digit
`datayear`), so a multi-year frame is labeled correctly:

```{r add-origin}
deaths <- data.frame(
    year          = c(2019L, 2019L, 2019L),
    age           = 25L,
    sex           = "male",
    race          = "white_only",
    hspanicr      = c(1, 6, 9),   # Hispanic, non-Hispanic, origin-unknown
    deaths        = c(40, 900, 5)
)
deaths <- add_hispanic_origin(deaths)
deaths[, c("hspanicr", "hispanic_origin", "deaths")]
```

To get Hispanic-stratified **rates**, put `hispanic_origin` in `by_vars` and use a
scheme that resolves origin -- `"single"` (2000+) or `"bridged"` (1990+). Drop
the origin-unknown deaths first: they have no denominator.

```{r stratified-rate}
strat <- deaths[deaths$hispanic_origin != "unknown", ]
add_pop_counts(
    strat,
    race_scheme = "single",
    by_vars = c("year", "age", "sex", "race", "hispanic_origin")
)[, c("hispanic_origin", "deaths", "pop")]
```

Each origin gets its own denominator. An `"unknown"` or `NA` origin left in the
frame is a hard error, not a silent all-origin fallback -- the counts for those
deaths belong in an all-origin numerator, never a stratified one.

When you carry this into an age-standardized rate (`add_std_pop()` then
`calc_stdrate_var()`), list `hispanic_origin` among the grouping variables you
pass to the rate helper, alongside the others (e.g. `year`, `sex`). Omitting a
stratifier does not error -- it silently averages over it -- so a forgotten
`hispanic_origin` returns one blended rate instead of a per-origin pair.

### Which knob? A shared name, two mechanisms

The join and the accessors both say `hispanic_origin`, but they are different
controls:

- `add_pop_counts()` joins on a `hispanic_origin` **column** you list in
  `by_vars` (above).
- `get_pop_state()` / `get_pop_county()` take a `hispanic_origin=` **filter
  argument** that selects which origin's population to return.

```{r accessor}
get_pop_state(scheme = "single", states = "06", years = 2024L,
              hispanic_origin = "hispanic")[1, ]
```

Confusing the two is loud, not silent: passing the filter argument to
`add_pop_counts()` errors (`unused argument`), and putting a `hispanic_origin`
column in a frame without listing it in `by_vars` errors on the stray column.

## The mixed-era trap

Note the race label changes with the scheme -- `bridged`/`legacy` use `white`,
`single` uses `white_only`; they are not interchangeable (this is exactly the
kind of quiet mismatch to avoid). SEER resolves Hispanic origin only from 1990.
Before 1990 the bridged denominators carry origin `"all"` only, so there is no
such thing as a pre-1990 Hispanic-specific rate -- the population does not
exist. `add_pop_counts()` enforces this: a pre-1990 row with a stratified
origin is a hard error.

That makes a continuous 1969-2024 "Hispanic trend" impossible, and narcan will
not let you fake one. A single frame that mixes the pre-1990 `"all"` marginal
with post-1990 stratified rows is rejected, because `"all"` already sums the
strata -- combining them double-counts:

```{r dd6-error}
mixed <- data.frame(
    year = c(1985L, 2000L), age = 40L, sex = "male", race = "white",
    hispanic_origin = c("all", "hispanic"), deaths = c(500, 30)
)
try(add_pop_counts(mixed, race_scheme = "bridged",
                   by_vars = c("year", "age", "sex", "race", "hispanic_origin")))
```

If you genuinely want to show the eras side by side, compute each **separately**
(each call internally homogeneous, so neither trips the guard) and `rbind()` the
results -- keeping the pre-1990 segment as all-origin:

```{r two-call, message = FALSE}
pre <- data.frame(year = 1985L, age = 40L, sex = "male", race = "white",
                  hispanic_origin = "all", deaths = 500)
pre <- add_pop_counts(pre, race_scheme = "bridged",
                      by_vars = c("year", "age", "sex", "race", "hispanic_origin"))

post <- expand.grid(year = c(1990L, 2000L), age = 40L, sex = "male",
                    race = "white",
                    hispanic_origin = c("hispanic", "non_hispanic"),
                    stringsAsFactors = FALSE)
post$deaths <- c(20, 480, 40, 520)
post <- add_pop_counts(post, race_scheme = "bridged",
                       by_vars = c("year", "age", "sex", "race", "hispanic_origin"))

trend <- rbind(pre, post)
trend$rate <- trend$deaths / trend$pop * 1e5
trend[order(trend$year), c("year", "hispanic_origin", "rate")]
```

The discontinuity is the point. The 1985 row is an all-origin rate; the 1990 and
2000 rows are Hispanic-specific. They are different series and must be read that
way -- the pre-1990 segment shows only the collapsed line, never a stratified
one. **Restrict origin-stratified trend analysis to 1990+ (bridged) or 2000+
(single);** the pre-1990 point exists here only to make the break visible.

## Two caveats for Hispanic-stratified rates

**Misclassification.** The numerator's Hispanic origin comes from the death
certificate; the denominator's comes from Census/SEER. They are measured
separately and are differentially misclassified -- modest for Hispanic and Asian
or Pacific Islander populations, large for American Indian or Alaska Native --
so origin-specific rates carry a numerator/denominator bias (Arias E, Heron M,
Hakes J. The Validity of Race and Hispanic-origin Reporting on Death Certificates
in the United States: An Update. *Vital Health Stat 2*(172). Hyattsville, MD:
National Center for Health Statistics; 2016).

**Incomplete early reporting.** The Hispanic-origin item was phased onto state
death certificates through about 1997, so national origin-stratified numerators
for 1990-1996 undercount Hispanic deaths and the resulting rates run low. The
undercount was concentrated in late-adopting states, so it is not a uniform
national adjustment. `add_pop_counts()` emits a once-per-session message when a
bridged join touches that span.

## See also

- `vignette("population-denominators")` -- the parent denominators topic:
  schemes, race labels, and the join guards this vignette builds on.
- `vignette("getting-started")` -- the package overview.
