---
title: "Population denominators: bridged, single-race, and legacy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population denominators: bridged, single-race, and legacy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A mortality rate is death counts over a matching population, and "matching" is
the whole problem. NCHS has coded race three incompatible ways over the series,
so narcan ships three denominator schemes and routes all of them through one
guarded join, `add_pop_counts()`. The schemes are NOT comparable -- you cannot
chain them into one trend -- so set `race_scheme` deliberately.

| Scheme | `race_scheme` | Source | Coverage | Use for |
|---|---|---|---|---|
| Legacy bridged | `"legacy"` (default) | frozen `pop_est` (single-race-alone PEP) | 1979-2020 | reproducing published bridged-race rates byte-for-byte |
| Single-race | `"single"` | Census PEP single-race (OMB 1997) | 2000-2024 | 2022+ single-race deaths (codes 101-106) |
| SEER bridged | `"bridged"` | SEER U.S. Population Data (Vintage 2024) | 1969-2024 | coherent bridged-race trends, including pre-2020 |

All data here are public (bundled Census/SEER estimates, or a small bundled
county fixture) plus synthetic death counts; no restricted NCHS records are
used, and no chunk downloads anything.

```{r libs}
library(narcan)
```

## `add_pop_counts()`, scheme by scheme

### Legacy (the default)

`race_scheme = "legacy"` joins the frozen `pop_est` and reproduces the
historical behavior byte-for-byte: unmatched keys warn and leave `pop = NA`. Use
it only to reproduce published bridged-race rates.

```{r legacy}
legacy <- data.frame(year = 1999L, age = 25L, sex = "male", race = "white")
add_pop_counts(legacy)                    # default scheme; adds `pop`
```

For 2000-2020 the legacy denominator is single-race-alone (it runs low against
bridged-race deaths), so `add_pop_counts()` nudges you once per session in that
span toward `"bridged"` or `"single"`.

### Single-race, end to end

`race_scheme = "single"` joins single-race denominators (2000-2024) for deaths
coded 101-106 (`white_only`/`black_only`/`american_indian_only`/`asian_only`/
`nhopi_only`/`multiracial`). It is strict: out-of-domain age/sex/race and any
unmatched key hard-error, so a denominator is never silently `NA`. Below is a
full `asian_only` age-standardized rate, one step at a time.

```{r single-deaths}
deaths <- expand.grid(
    year = 2024L, age = seq(0, 85, 5), sex = c("male", "female"),
    race = "asian_only", stringsAsFactors = FALSE
)
deaths$deaths <- rep(c(1, 2, 5, 12, 30, 55), length.out = nrow(deaths))
head(deaths)
```

`add_pop_counts()` adds the matched `pop` column:

```{r single-pop}
rated <- add_pop_counts(deaths, race_scheme = "single")
head(rated)
```

`add_std_pop()` adds the US 2000 standard population (`pop_std`) and its unit
weights (`unit_w`, summing to 1 across age groups):

```{r single-std-pop}
weighted <- add_std_pop(rated)
head(weighted[, c("age", "sex", "pop", "pop_std", "unit_w")])
```

`calc_asrate_var()` adds the **age-specific** rate per 100,000 (`drug_rate`) and
its Poisson variance (`drug_var`) -- one rate per age/sex cell:

```{r single-asrate}
asr <- calc_asrate_var(weighted, new_name = drug, death_col = deaths)
head(asr[, c("age", "sex", "deaths", "pop", "drug_rate", "drug_var")])
```

`calc_stdrate_var()` collapses the age bins into one **age-standardized** rate
per group (grouping is not automatic -- pass every dimension to keep):

```{r single-stdrate}
calc_stdrate_var(asr, drug_rate, drug_var, sex, race)
```

### Bridged (SEER-uniform, era-ragged)

`race_scheme = "bridged"` joins the SEER bridged denominators (1969-2024). It is
strict and era-ragged: SEER resolves AIAN/API and Hispanic origin only from 1990
(pre-1990 is white/black/other), so `year` MUST be a join key.

```{r bridged}
api <- data.frame(year = 2019L, age = 40L, sex = "female", race = "api")
add_pop_counts(api, race_scheme = "bridged",
               by_vars = c("year", "age", "sex", "race"))
```

Omitting `year` from `by_vars` is an error -- the valid race set is
era-dependent:

```{r bridged-noyear, error = TRUE}
add_pop_counts(api, race_scheme = "bridged",
               by_vars = c("age", "sex", "race"))
```

And an `api` row before 1990 errors, because SEER has no AIAN/API split yet:

```{r bridged-pre1990, error = TRUE}
old_api <- data.frame(year = 1985L, age = 40L, sex = "female", race = "api")
add_pop_counts(old_api, race_scheme = "bridged",
               by_vars = c("year", "age", "sex", "race"))
```

## Strictness guards

Legacy and bridged share the labels white/black/other/total, so a mis-set
`race_scheme` cannot always be detected. But single-race labels under a
non-single scheme always can be -- passing them to the default is a hard error,
a guard against forgetting `race_scheme = "single"`:

```{r contradiction, error = TRUE}
bad <- data.frame(year = 2024L, age = 25L, sex = "male", race = "asian_only")
add_pop_counts(bad)                       # default is "legacy" -> error
```

A frame already summed over a dimension uses a reserved token -- `race = "total"`,
`sex = "both"`, `hispanic_origin = "all"`. The matching marginal is synthesized
on demand (finest cells summed; no marginal is stored, so it cannot double-count):

```{r tokens}
both_sex <- data.frame(year = 2024L, age = 65L, sex = "both", race = "asian_only")
add_pop_counts(both_sex, race_scheme = "single")$pop        # male + female
```

## State and county grains

Include `state_fips` (or `county_fips`) in `by_vars` to route to sub-national
denominators. The state single-race table is bundled, so a state join needs no
extra package:

```{r state}
ca <- data.frame(state_fips = "06", year = 2024L, age = 40L, sex = "female",
                 race = "asian_only", deaths = 20)
add_pop_counts(ca, race_scheme = "single",
               by_vars = c("state_fips", "year", "age", "sex", "race"))
```

A geography column that is in the frame but absent from `by_vars` is an error,
never a silent national join:

```{r stray, error = TRUE}
add_pop_counts(ca, race_scheme = "single",
               by_vars = c("year", "age", "sex", "race"))
```

Public MCOD carries state/county geographic detail only through data year 2004
(2005+ county detail needs restricted NCHS files). The county denominators are
distributed as a downloadable parquet; here we point narcan at the small bundled
Wyoming fixture instead of a live download.

```{r county, eval = requireNamespace("duckdb", quietly = TRUE)}
fx <- system.file("extdata", "pop_singlerace_county_fixture.parquet",
                  package = "narcan")

wy_deaths <- data.frame(
    county_fips = "56013", year = 1999:2004, age = 40L, sex = "male",
    race = "white_only", deaths = c(1, 2, 4, 3, 5, 2)
)
wy_deaths

options(narcan.pop_single_county_parquet = fx)   # normally a downloaded parquet
add_pop_counts(subset(wy_deaths, year >= 2000), race_scheme = "single",
               by_vars = c("county_fips", "year", "age", "sex", "race"))
```

Single-race denominators start in 2000, so the 1999 row is dropped before the
join above -- death-side coverage (public MCOD geography) and denominator-side
coverage (single-race) can differ, so join only where both exist.

## Coverage and the frozen slice

The single-race 2020-2024 slice is frozen at 0.5.0; the 2000-2024 backfill is
additive and does not change those five years. A request for a year the resolved
data does not cover hard-errors -- never a silent 0-row or `NA`:

```{r coverage, error = TRUE}
get_pop_state(scheme = "single", years = 2025)
```

## Provenance

`pop_sources()` prints the manifest -- scheme, grain, vintage, coverage, and
delivery for every dataset. Check a bundled dataset's vintage before mixing it
with a freshly downloaded file.

```{r sources}
pop_sources()
```

## A note on hand-joins

Always join through `add_pop_counts()` (or the guarded rate helpers). The
descriptive accessors `get_pop_state()`/`get_pop_county()` return population rows
but do NOT guard a hand-join -- if you `left_join()` a slice yourself, or query
the parquet with raw SQL, you own verifying that the slice is unique on your keys
and that no key is silently dropped. The guarded path exists precisely to make
those mistakes impossible.
