---
title: "Population denominators: bridged, single-race, and legacy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population denominators: bridged, single-race, and legacy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A mortality rate is death counts over a matching population, and "matching" is
the whole problem. NCHS has coded race three incompatible ways over the series,
so narcan ships three denominator schemes and routes all of them through one
guarded join, `add_pop_counts()`. The schemes are NOT comparable -- you cannot
chain them into one trend -- so set `race_scheme` deliberately.

| Scheme | `race_scheme` | Source | Coverage | Use for |
|---|---|---|---|---|
| Legacy (historical NCHS) | `"legacy"` (default) | frozen `pop_est` (single-race-alone PEP) | 1979-2020 | reproducing prior narcan-based analyses byte-for-byte |
| Single-race | `"single"` | Census PEP single-race (OMB 1997) | 2000-2024 | 2021+ single-race deaths (codes 101-106) |
| SEER bridged | `"bridged"` | SEER U.S. Population Data (Vintage 2024) | 1969-2024 (population) | coherent bridged-race trends, deaths through 2020 |

(PEP = Census Population Estimates Program; OMB 1997 = the 1997 OMB race/ethnicity
standards; SEER = NCI's Surveillance, Epidemiology, and End Results program.)

All data here are public (bundled Census/SEER estimates, or a small bundled
county fixture) plus synthetic death counts; no restricted NCHS records are
used, and no chunk downloads anything.

```{r libs}
library(narcan)
```

Across every scheme, `add_pop_counts()` joins on `by_vars` -- default
`c("year", "age", "sex", "race")`, which returns a national denominator; adding
`state_fips` or `county_fips` to `by_vars` routes the same call to sub-national
denominators.

## `add_pop_counts()`, scheme by scheme

### Legacy (the default)

`race_scheme = "legacy"` joins the frozen `pop_est` and reproduces narcan's
historical behavior byte-for-byte: unmatched keys warn and leave `pop = NA`.
`pop_est` is a pieced-together series whose race labels are
`white`/`black`/`other`/`nhw` (not the four bridged categories) and whose
2000-2020 denominators are single-race-alone Census estimates. It exists to
reproduce **prior narcan-based analyses** exactly -- it is not the official NCHS
bridged-race series. For a coherent bridged-race denominator use `"bridged"`; for
2000+ single-race deaths use `"single"`.

```{r legacy}
legacy <- data.frame(year = 1999L, age = 25L, sex = "male", race = "white")
add_pop_counts(legacy)                    # default scheme; adds `pop`
```

For 2000-2020 the legacy denominator is single-race-alone, which undercounts the
multiple-race population and so inflates race-specific rates (especially for the
AIAN and API groups); `add_pop_counts()` nudges you once per session in that span
toward `"bridged"` or `"single"`.

### Single-race

`race_scheme = "single"` joins single-race denominators (2000-2024) for deaths
coded 101-106 (`white_only`/`black_only`/`american_indian_only`/`asian_only`/
`nhopi_only`/`multiracial`; `nhopi` = Native Hawaiian or Other Pacific Islander).
It is strict: out-of-domain age/sex/race and any unmatched key hard-error, so a
denominator is never silently `NA`.

```{r single-deaths}
deaths <- expand.grid(
    year = 2024L, age = seq(0, 85, 5), sex = c("male", "female"),
    race = "asian_only", stringsAsFactors = FALSE
)
deaths$deaths <- rep(c(1, 2, 5, 12, 30, 55), length.out = nrow(deaths))
head(deaths)
```

`add_pop_counts()` adds the matched `pop` column:

```{r single-pop}
rated <- add_pop_counts(deaths, race_scheme = "single")
head(rated)
```

From the matched `pop`, the rate pipeline (`add_std_pop()` ->
`calc_asrate_var()` -> `calc_stdrate_var()`) turns these counts into an
age-standardized rate. That pipeline -- including the standard-population choice
and confidence intervals -- is walked through step by step in
`vignette("age-standardized-rates")`.

### Bridged (SEER-uniform, era-ragged)

`race_scheme = "bridged"` joins the SEER bridged denominators (1969-2024). It is
strict and era-ragged (the set of valid race and origin categories changes by
year, so coverage is uneven across eras): SEER resolves AIAN/API (American Indian
or Alaska Native / Asian or Pacific Islander) and Hispanic origin only from 1990
(pre-1990 is white/black/other), so `year` MUST be a join key.

The "1969-2024" coverage is *population* availability. On the death side, NCHS
bridged-race coding ends with data year 2020 (2021 is reserved; single-race codes
101-106 begin in 2022), so there is no bridged-race-coded death series to pair
with the 2021-2024 population -- use `"single"` for 2021+ deaths. Treat `"bridged"`
as a coherent race series **through 2020**.

```{r bridged}
api <- data.frame(year = 2019L, age = 40L, sex = "female", race = "api")
add_pop_counts(api, race_scheme = "bridged",
               by_vars = c("year", "age", "sex", "race"))
```

Omitting `year` from `by_vars` is an error -- the valid race set is
era-dependent:

```{r bridged-noyear, error = TRUE}
add_pop_counts(api, race_scheme = "bridged",
               by_vars = c("age", "sex", "race"))
```

And an `api` row before 1990 errors, because SEER has no AIAN/API split yet:

```{r bridged-pre1990, error = TRUE}
old_api <- data.frame(year = 1985L, age = 40L, sex = "female", race = "api")
add_pop_counts(old_api, race_scheme = "bridged",
               by_vars = c("year", "age", "sex", "race"))
```

## Strictness guards

Legacy and bridged share the labels white/black/other/total, so a mis-set
`race_scheme` cannot always be detected. But single-race labels under a
non-single scheme always can be -- passing them to the default is a hard error,
a guard against forgetting `race_scheme = "single"`:

```{r contradiction, error = TRUE}
bad <- data.frame(year = 2024L, age = 25L, sex = "male", race = "asian_only")
add_pop_counts(bad)                       # default is "legacy" -> error
```

A frame already summed over a dimension uses a reserved token -- `race = "total"`,
`sex = "both"`, `hispanic_origin = "all"`. The matching marginal is synthesized
on demand (finest cells summed; no marginal is stored, so it cannot double-count):

```{r tokens}
both_sex <- data.frame(year = 2024L, age = 65L, sex = "both", race = "asian_only")
add_pop_counts(both_sex, race_scheme = "single")$pop        # male + female
```

## State and county grains

Include `state_fips` (or `county_fips`) in `by_vars` to route to sub-national
denominators. The state single-race table is bundled, so a state join needs no
extra package. See `vignette("geography-fips")` for how to derive harmonized
`state_fips`/`county_fips` columns from raw NCHS fields.

```{r state}
ca <- data.frame(state_fips = "06", year = 2024L, age = 40L, sex = "female",
                 race = "asian_only", deaths = 20)
add_pop_counts(ca, race_scheme = "single",
               by_vars = c("state_fips", "year", "age", "sex", "race"))
```

A geography column that is in the frame but absent from `by_vars` is an error,
never a silent national join:

```{r stray, error = TRUE}
add_pop_counts(ca, race_scheme = "single",
               by_vars = c("year", "age", "sex", "race"))
```

Public MCOD carries state/county geographic detail only through data year 2004
(2005+ county detail needs restricted NCHS files). The county denominators are
distributed as a downloadable parquet; here we point narcan at the small bundled
Wyoming fixture instead of a live download.

This example needs the optional `duckdb` package (`install.packages("duckdb")`);
with it absent, the chunk below simply produces no output.

```{r county, eval = requireNamespace("duckdb", quietly = TRUE)}
fx <- system.file("extdata", "pop_singlerace_county_fixture.parquet",
                  package = "narcan")

wy_deaths <- data.frame(
    county_fips = "56013", year = 1999:2004, age = 40L, sex = "male",
    race = "white_only", deaths = c(1, 2, 4, 3, 5, 2)
)
wy_deaths

options(narcan.pop_single_county_parquet = fx)   # normally a downloaded parquet
add_pop_counts(subset(wy_deaths, year >= 2000), race_scheme = "single",
               by_vars = c("county_fips", "year", "age", "sex", "race"))
```

Single-race denominators start in 2000, so the 1999 row is dropped before the
join above -- death-side coverage (public MCOD geography) and denominator-side
coverage (single-race) can differ, so join only where both exist.

## Coverage and the frozen slice

The **national** single-race table is bundled at every covered year: the frozen
0.5.0 `pop_singlerace` (2020-2024) plus `pop_singlerace_full` (2000-2024,
dependency-free `.rda`), so a national pre-2020 request needs no download. The
2000-2019 backfill is additive -- it does not change those five years. The
**state and county** backfills are the ones distributed as downloadable
single-race `*_full` parquets (tag-pinned GitHub Release assets, cached on first
use). So `get_pop_state()`/`get_pop_county()` serve 2020-2024 from the bundled
frozen table, but a pre-2020 `years` request at state/county grain routes to that
downloadable parquet -- or to a local copy via the
`narcan.pop_single_<grain>_parquet` option, as in the county chunk above. A
request for a year the resolved data does not cover hard-errors -- never a silent
0-row or `NA`:

```{r coverage, error = TRUE}
get_pop_state(scheme = "single", years = 2025)
```

## Provenance

`pop_sources()` prints the manifest -- dataset, scheme, grain, vintage, and
coverage for every single-race and bridged dataset (the frozen legacy `pop_est`
predates this download-manifest system). Check a bundled dataset's vintage before
mixing it with a freshly downloaded file.

```{r sources, results = "asis"}
invisible(utils::capture.output(src <- pop_sources()))
knitr::kable(src[, c("dataset", "scheme", "grain", "vintage",
                     "year_min", "year_max")])
```

## Hispanic-stratified denominators

Both strict schemes carry a Hispanic-origin axis. For an origin-specific
denominator, add a `hispanic_origin` column (from `add_hispanic_origin()`) to
`by_vars`; for the all-origin denominator, drop the column entirely. The two
answer different questions -- and origin-unknown deaths belong only in the
all-origin numerator, because there is no "unknown-origin" population to divide
by.

```{r hispanic-all}
# All-origin: one row, every death counts (the 3 origin-unknown deaths are here).
all_origin <- data.frame(year = 2024L, age = 30L, sex = "female",
                         race = "white_only", deaths = 12 + 210 + 3)
add_pop_counts(all_origin, race_scheme = "single",
               by_vars = c("year", "age", "sex", "race"))$pop
```

```{r hispanic-strat}
# Origin-stratified: the non-denominable unknowns are dropped; origin is a key.
# ('denominable' means it has a matching population to divide by.)
strat <- data.frame(
    year = 2024L, age = 30L, sex = "female", race = "white_only",
    hispanic_origin = c("hispanic", "non_hispanic"), deaths = c(12, 210)
)
add_pop_counts(strat, race_scheme = "single",
               by_vars = c("year", "age", "sex", "race", "hispanic_origin")
)[, c("hispanic_origin", "deaths", "pop")]
```

The two stratified denominators sum to the all-origin one; the difference is only
which deaths sit in the numerator. Leaving a `hispanic_origin` column in the frame
but out of `by_vars` is a hard error (it would be silently summed over), as is an
`"unknown"`/`NA` origin in a stratified join. The detailed-vs-binary recode
distinction, the pre-1990 mixed-era trap, and the misclassification caveats are
covered in `vignette("hispanic-origin")`.

## A note on hand-joins

Always join through `add_pop_counts()` (or the guarded rate helpers). The
descriptive accessors `get_pop_state()`/`get_pop_county()` return population rows
but do NOT guard a hand-join -- if you `left_join()` a slice yourself, or query
the parquet with raw SQL, you own verifying that the slice is unique on your keys
and that no key is silently dropped. The guarded path exists precisely to make
those mistakes impossible.

## See also

- `vignette("getting-started")` -- the package overview and where population
  denominators fit in the workflow.
- `vignette("age-standardized-rates")` -- feeds these denominators through a full
  age-standardized rate pipeline end to end.
