---
title: "End-to-end on real public-use data (2004)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end on real public-use data (2004)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The other vignettes use small synthetic frames to keep each step legible. This one
runs the whole pipeline on a **real** file, start to finish, and reports the actual
numbers -- from raw fixed-width records to an age-standardized opioid death rate by
sex, for 2004.

The file is the NCHS **public-use** Multiple Cause of Death (MCOD) file for 2004, so
no restricted-data agreement is needed. 2004 is the last public-use year that still
carries county geography (2005 onward suppress it), which makes it a DUA-free
stand-in for developing county-level code before you obtain the restricted
All-County files -- see the note on `import_mcod_fwf()` in
`vignette("getting-started")`.

```{r libs}
library(narcan)
```

The heavy steps -- reading the ~1.2 GB fixed-width file and flagging all 2.4 million
records -- are shown with `eval = FALSE`, because the file is not bundled with the
package and cannot be downloaded during a package check. The counts they produce are
reproduced below as a small table, so the rate half of the pipeline (the part that
uses only narcan's bundled population data) **runs live** on the real 2004 counts.

## Get the file

Download the public-use MCOD fixed-width file for 2004 (distributed by NCHS and
mirrored by NBER as `mort2004us.zip`). Unzip it and pass the extracted `.dat` to
`import_mcod_fwf()`, which reads the raw fixed-width layout -- the archive's internal
filename may differ from the placeholder below, so point the call at whatever the
unzip produces. narcan's `download_mcod_csv()` / `download_mcod_dta()` fetch the
CSV/Stata versions instead, which are a different on-ramp not used here.

```{r import, eval = FALSE}
raw <- import_mcod_fwf("mort2004us.dat", year = 2004, tier = "public")
nrow(raw)
#> [1] 2401400
```

`import_mcod_fwf()` returns every column in the restricted layout order; for 2004 the
public and restricted tiers are effectively identical (the one restricted-only field,
`racer40`, sits past the end of the public record and comes back all-`NA`).

## Flag the records

Keep US residents, collapse the 20 contributory-cause columns into `f_records_all`,
then run the ISW7 flag pipeline. Each step keys on `year = 2004` to select the ICD-10
coding era.

```{r flag, eval = FALSE}
flagged <- raw |>
    subset_residents() |>                    # restatus 1-3; drops non-residents
    unite_records(year = 2004) |>            # builds f_records_all
    flag_drug_deaths(year = 2004) |>
    flag_opioid_deaths(year = 2004) |>
    flag_opioid_types(year = 2004)
```

Here are the national totals that pipeline produces on the real 2004 file:

```{r totals}
totals <- data.frame(
    quantity = c("US resident deaths", "drug deaths (ISW7)",
                 "opioid deaths (ISW7)", "heroin, T40.1", "synthetic, T40.4"),
    n = c(2397615L, 27393L, 13756L, 1878L, 1664L)
)
totals
```

narcan's `drug_death` is the ISW7 multiple-cause definition -- a drug-poisoning
underlying cause **and** a drug T-code -- so it runs a little below the CDC WONDER
"drug overdose" headline, which keys on the underlying cause alone. The two are close
but not identical; see `vignette("classifying-overdose-deaths")` for the exact rules.

## Recode age and sex

Standardizing by age needs age in the standard population's 5-year bins, and the
population join needs `sex` as `"male"`/`"female"`. Sex coding changed at data year
2003 (numeric `1`/`2` before, `"M"`/`"F"` after), so `categorize_sex()` takes `year`.
`convert_ager27()` needs no year -- the 1-27 AGER27 scheme is stable across
eras; `vignette("demographic-recodes")` covers both in full.

```{r recode, eval = FALSE}
flagged$sex <- categorize_sex(flagged$sex, year = 2004)   # "M"/"F" -> male/female
flagged <- convert_ager27(flagged)                        # ager27 -> 5-year `age`
```

## Aggregate to counts

Collapse the flagged records to opioid-death counts by `year x age x sex`. In a real
script this is the line that turns 2.4 million records into the 36-row table below.

```{r aggregate, eval = FALSE}
agg <- flagged |>
    dplyr::filter(!is.na(age), !is.na(sex)) |>
    dplyr::group_by(year, age, sex) |>
    dplyr::summarize(opioid_deaths = sum(opioid_death), .groups = "drop")
```

The `!is.na()` filter drops the four opioid deaths with an unstated age (none had an
unstated sex in this file), so the stratified table sums to 13,752 -- four short of the
13,756 national total above. Deaths with a missing stratifier cannot be placed in an
age-sex cell.

That aggregate is small enough to reproduce verbatim -- these are the **real 2004
opioid-death counts** by 5-year age group and sex:

```{r counts}
agg <- tibble::tribble(
    ~age, ~sex,       ~opioid_deaths,
      0L, "female",     11L,   5L, "female",      1L,  10L, "female",     12L,
     15L, "female",     86L,  20L, "female",    247L,  25L, "female",    301L,
     30L, "female",    408L,  35L, "female",    587L,  40L, "female",    910L,
     45L, "female",    910L,  50L, "female",    569L,  55L, "female",    293L,
     60L, "female",    139L,  65L, "female",     63L,  70L, "female",     36L,
     75L, "female",     35L,  80L, "female",     19L,  85L, "female",     14L,
      0L, "male",       15L,   5L, "male",        1L,  10L, "male",       16L,
     15L, "male",      368L,  20L, "male",      895L,  25L, "male",      948L,
     30L, "male",      916L,  35L, "male",     1139L,  40L, "male",     1593L,
     45L, "male",     1538L,  50L, "male",      984L,  55L, "male",      432L,
     60L, "male",      141L,  65L, "male",       60L,  70L, "male",       27L,
     75L, "male",       16L,  80L, "male",        9L,  85L, "male",       13L
)
agg$year <- 2004L
```

One subtlety matters for standardization. The standard population spans all 18 age
groups, so every age-sex cell needs its denominator even when it saw zero opioid
deaths -- otherwise the age standard is silently truncated. (In 2004 every cell
happens to have at least one opioid death, but this step is what protects the rate
when one does not.) Build the full grid and fill any empty cells with 0:

```{r grid}
counts <- expand.grid(year = 2004L, age = seq(0L, 85L, 5L),
                      sex = c("male", "female"), stringsAsFactors = FALSE)
counts <- merge(counts, agg, all.x = TRUE)
counts$opioid_deaths[is.na(counts$opioid_deaths)] <- 0L
nrow(counts)   # 36 = 18 age groups x 2 sexes
```

## Denominators and the rate

From here it is the standard rate pipeline (see `vignette("age-standardized-rates")`),
running live on the real counts. Use the SEER bridged-race denominators
(`race_scheme = "bridged"`), which cover 2004 as one internally consistent series;
race is not a `by_var`, so `pop` is the all-race total per `year x age x sex`.

```{r pop}
counts <- add_pop_counts(counts, by_vars = c("year", "age", "sex"),
                         race_scheme = "bridged")
counts <- add_std_pop(counts, std_cat = "s204", by_vars = "age")
counts <- calc_asrate_var(counts, new_name = opioid,
                          death_col = opioid_deaths, pop_col = pop)
head(counts[order(counts$age, counts$sex),
            c("age", "sex", "opioid_deaths", "pop", "opioid_rate")])
```

`calc_stdrate_var()` collapses the 18 age bins into one age-standardized rate per
sex, reweighting to the US 2000 standard. A 95% Wald interval follows from the
returned variance.

```{r stdrate}
std <- calc_stdrate_var(counts, opioid_rate, opioid_var, year, sex)
std <- as.data.frame(std)
std$lower <- std$opioid_rate - 1.96 * sqrt(std$opioid_var)
std$upper <- std$opioid_rate + 1.96 * sqrt(std$opioid_var)
std[, c("year", "sex", "opioid_rate", "lower", "upper")]
```

In 2004 the age-standardized opioid death rate was about **6.2 per 100,000 among men
and 3.1 among women** -- roughly a two-to-one gap, standardized to the same age
structure, so the difference is in death rates and not in age composition. These are
the real figures the pipeline recovers from the public-use file, computed live in the
chunk above.

```{r plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.alt = "Bar chart of the age-standardized opioid death rate per 100,000 in 2004 by sex. The male rate is about 6.2 and the female rate about 3.1, with narrow 95% confidence-interval bars.", fig.cap = "Age-standardized opioid death rate per 100,000 by sex, 2004 public-use MCOD (US 2000 standard). Bars are 95% confidence intervals."}
library(ggplot2)

ggplot(std, aes(x = sex, y = opioid_rate, fill = sex)) +
    geom_col(width = 0.6) +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
    labs(x = NULL, y = "Age-standardized rate per 100,000",
         title = "Opioid death rate by sex, 2004") +
    theme_minimal() +
    theme(legend.position = "none")
```

## Caveats

**County geography on the public file.** County FIPS are populated only for counties
with a 2000-Census population of at least 100,000; smaller (disproportionately rural)
counties collapse to a residual code. The 2004 stand-in exercises large-county
behavior only -- the restricted All-County files carry every county.

**Definition.** These counts use narcan's ISW7 multiple-cause definitions
(`vignette("classifying-overdose-deaths")`), which differ modestly from the
underlying-cause-only counts CDC WONDER reports.

**Small-count intervals.** The Wald interval above is fine for these well-populated
national cells but under-covers small strata; use a gamma-based interval (Fay and
Feuer, *Stat Med* 1997) for sparse sub-national or fine demographic cells.

**Cross-era comparability.** A 2004 rate is directly comparable to other ICD-10 years
(1999+), but not across the 1999 ICD-9-to-ICD-10 revision without a comparability
adjustment, nor across a change in the denominator race scheme.

## See also

- `vignette("getting-started")` -- the package overview and the import on-ramp.
- `vignette("classifying-overdose-deaths")` -- the ISW7 flag rules used here.
- `vignette("population-denominators")` -- the denominator schemes and the join used
  in the rate step.
- `vignette("age-standardized-rates")` -- the rate pipeline in more detail.
- `vignette("demographic-recodes")` -- the age/sex/race recoders across eras.
- `vignette("geography-fips")` -- more on the county-FIPS caveat noted above.
