---
title: "Single-race mortality rates (2020-2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-race mortality rates (2020-2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

From data year 2022, NCHS reports **single-race** (OMB 1997) categories instead
of the bridged-race scheme used through 2020. `remap_race()`/`categorize_race()`
map those deaths to codes 101-106 (`white_only`, `black_only`,
`american_indian_only`, `asian_only`, `nhopi_only`, `multiracial`). Rates need
matching single-race **denominators** -- dividing single-race death counts by the
bridged-race `pop_est` would be wrong.

This vignette computes an `asian_only` age-standardized death rate end to end. All
data here are public Census population estimates (bundled or a small bundled
county fixture) plus a synthetic death count; no restricted NCHS records are
used, and no chunk downloads anything.

```{r libs}
library(narcan)
```

## Two denominator schemes -- never mix them

`add_pop_counts()` takes a `race_scheme`:

- `"legacy"` (the default) joins the frozen bridged-race `pop_est` (1979-2020)
  and reproduces published bridged-race rates byte-for-byte.
- `"single"` joins the single-race denominators (2020-2024) for 101-106 deaths.

The single scheme is strict: out-of-domain `age`/`sex`/`race` values and any
unmatched key are hard errors, so a denominator is never silently `NA`. Passing
single-race labels under the default scheme is itself an error -- a guard against
an incorrectly set `race_scheme`:

```{r contradiction, error = TRUE}
bad <- data.frame(year = 2024, age = 25, sex = "male", race = "asian_only")
add_pop_counts(bad)                       # default is "legacy" -> error
```

## National `asian_only` rate

Build a small aggregated death frame (the counts here are illustrative) and join
national single-race denominators. The `race` column holds `categorize_race()`
labels; `age` is the 5-year bin start.

```{r national}
deaths <- expand.grid(
    year = 2024L, age = seq(0, 85, 5), sex = c("male", "female"),
    race = "asian_only", stringsAsFactors = FALSE
)
deaths$deaths <- rep(c(1, 2, 5, 12, 30, 55), length.out = nrow(deaths))

rated <- add_pop_counts(deaths, race_scheme = "single")
head(rated)
```

Age-standardize to the US 2000 standard with the existing rate helpers:

```{r std}
asr <- rated |>
    add_std_pop() |>
    calc_asrate_var(new_name = drug, death_col = deaths)
head(asr[, c("age", "sex", "pop", "drug_rate")])
```

## Reserved aggregate tokens

A death frame that is already summed over a dimension uses a reserved token --
`race = "total"`, `sex = "both"`, `hispanic_origin = "all"`. The matching
population marginal is synthesized on demand (finest cells are summed; no
marginal is ever stored, so aggregation cannot double-count).

```{r tokens}
both_sex <- data.frame(year = 2024L, age = 65L, sex = "both", race = "asian_only")
add_pop_counts(both_sex, race_scheme = "single")$pop        # male + female
```

## State and county grains

Include `state_fips` (or `county_fips`) in `by_vars` to route to sub-national
denominators. A frame that carries a geography column absent from `by_vars` is an
error, never a silent national join.

```{r state}
ca <- data.frame(state_fips = "06", year = 2024L, age = 40L, sex = "female",
                 race = "asian_only", deaths = 20)
add_pop_counts(ca, race_scheme = "single",
               by_vars = c("state_fips", "year", "age", "sex", "race"))
```

The county grain is distributed as a downloadable parquet (see
`download_pop_data()` and `pop_sources()`); here we read the small bundled
Wyoming fixture directly. Sparse single-race county cells are legitimately zero,
which gives an undefined rate -- aggregate ages or geographies for a defined one.

```{r county, eval = requireNamespace("duckdb", quietly = TRUE)}
fx <- system.file("extdata", "pop_singlerace_county_fixture.parquet",
                  package = "narcan")
wy_asian <- get_pop_county(states = "56", years = 2024L, parquet = fx)
wy_asian <- wy_asian[wy_asian$race == "asian_only", ]
sum(wy_asian$pop == 0)          # many zero-pop asian_only county-age-sex cells
```

## Provenance

`pop_sources()` prints the manifest -- scheme, grain, vintage, coverage, and how
each dataset is delivered. Check the bundled datasets' vintage before mixing them
with a freshly downloaded county file.

```{r sources}
pop_sources()
```

## A note on hand-joins

Always join through `add_pop_counts()` (or the guarded rate helpers). If you hand
`left_join()` a population slice yourself -- or query the parquet with raw SQL --
you get **unguarded** rows: you are then responsible for verifying the slice is
unique on your join keys and that no key is silently dropped. The guarded path
exists precisely to make those mistakes impossible.
