---
title: "Trends in unspecified drug overdose deaths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trends in unspecified drug overdose deaths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Many drug-overdose death certificates never name the specific drug involved. That
gap biases drug-specific trends -- an apparent rise in one opioid can reflect
improved toxicology and reporting specificity rather than a real increase in that
drug. In ICD-10 data (1999 onward), the opioid deaths with no specific type are
coded T40.6, "other and unspecified narcotics", which narcan flags as
`other_op_present`. This vignette measures the T40.6 share of opioid deaths by
year, end to end, on small illustrative records; no restricted NCHS data are used
and no chunk downloads anything.

```{r libs}
library(narcan)
```

## Illustrative records

The frame below is **synthetic and illustrative**. Each row is one death, coded
in the ICD-10 convention -- `ucod` is the underlying-cause code and
`f_records_all` is the space-joined multiple-cause T-code string. Opioid deaths
carry a poisoning `ucod` (X42, accidental poisoning by narcotics) plus one T40.x
opioid code: heroin (T40.1), synthetic opioids such as fentanyl (T40.4), or the
"other and unspecified" residual (T40.6). Two cocaine deaths per year (T40.5) are
included so the opioid filter has something to remove. Counts are chosen so the
T40.6 share falls as more deaths get a specific opioid type over time.

```{r data}
recipe <- tibble::tribble(
    ~year, ~ucod, ~f_records_all, ~n,
    1999L, "X42", "T401",         4L,   # heroin
    1999L, "X42", "T404",         2L,   # synthetic (e.g. fentanyl)
    1999L, "X42", "T406",         4L,   # other/unspecified opioid
    1999L, "X42", "T405",         2L,   # cocaine (non-opioid drug death)
    2005L, "X42", "T401",         5L,
    2005L, "X42", "T404",         3L,
    2005L, "X42", "T406",         4L,
    2005L, "X42", "T405",         2L,
    2012L, "X42", "T401",         6L,
    2012L, "X42", "T404",         6L,
    2012L, "X42", "T406",         3L,
    2012L, "X42", "T405",         2L,
    2019L, "X42", "T401",         8L,
    2019L, "X42", "T404",         9L,
    2019L, "X42", "T406",         3L,
    2019L, "X42", "T405",         2L
)
deaths <- recipe[rep(seq_len(nrow(recipe)), recipe$n),
                 c("year", "ucod", "f_records_all")]
rownames(deaths) <- NULL
nrow(deaths)
```

## Walk the flags, one step at a time

Each flag keys on `ucod` plus `f_records_all`, and the ICD era is read from
`year`, so run the pipeline one year at a time. Here is the 2012 slice through
each step.

`flag_drug_deaths()` adds `drug_death` -- for ICD-10 it is 1 when a qualifying
poisoning `ucod` pairs with a drug T-code. All four 2012 record types qualify.

```{r flag_drug}
d12 <- deaths[deaths$year == 2012, ]
s1 <- flag_drug_deaths(d12, year = 2012)
head(s1)
```

`flag_opioid_deaths()` adds `opioid_death`, which additionally requires a T40.0-.4
or T40.6 opioid code. The three opioid rows stay 1; the cocaine death (T40.5)
drops to 0.

```{r flag_opioid}
s2 <- flag_opioid_deaths(s1, year = 2012)
s2[c(1, 13, 16), ]             # a heroin, an other/unspecified, a cocaine row
```

`flag_opioid_types()` adds the six specific-type columns, the ICD-9-era residual
`unspecified_op_present`, and `num_opioids`. The T40.6 deaths land in
`other_op_present`; `unspecified_op_present` stays 0 (see the closing note).

```{r flag_types}
s3 <- flag_opioid_types(s2, year = 2012)
type_cols <- c("ucod", "f_records_all", "opioid_death", "heroin_present",
               "other_synth_present", "other_op_present",
               "unspecified_op_present", "num_opioids")
unique(s3[s3$opioid_death == 1, type_cols])
```

## Unspecified (T40.6) share by year

Apply the same three steps to every year, keep the opioid deaths, and take the
mean of `other_op_present` per year -- that mean is the T40.6 share.

```{r aggregate}
flagged <- dplyr::bind_rows(lapply(split(deaths, deaths$year), function(d) {
    y <- d$year[1]
    d |>
        flag_drug_deaths(year = y) |>
        flag_opioid_deaths(year = y) |>
        flag_opioid_types(year = y)
}))

unspec <- flagged |>
    dplyr::filter(opioid_death == 1) |>
    dplyr::group_by(year) |>
    dplyr::summarize(
        opioid_deaths = dplyr::n(),
        t406_unspecified = sum(other_op_present),
        t406_share = round(mean(other_op_present), 3),
        .groups = "drop"
    )
unspec
```

The T40.6 share falls from 0.40 in 1999 to 0.15 in 2019 -- built into these
illustrative counts, but the identical computation on real MCOD records recovers
whatever trend the data actually hold.

## Caveats

T40.6 is the standard ICD-10 proxy for an unspecified opioid, but the label is
"other **and** unspecified narcotics", so a T40.6 death is one where the specific
opioid was not distinguished, not strictly one that is missing. This is the
opioid *type*-unspecified share, not the broader problem of drug deaths with no
specific drug coded at all, which spans non-opioid drugs too. narcan also carries
`unspecified_op_present`, the ICD-9-era residual keyed to the generic opiate code
965.0; it is 0 by construction for years >= 1999 because ICD-10 folds the
unspecified opioids into T40.6, which is exactly why the modern measure reads
`other_op_present` instead. Some analysts redistribute T40.6 deaths across the
known types before computing drug-specific rates; narcan gives you the raw flags
so you can measure the unspecified share first and decide how to adjust for it
yourself.
